How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-04-11 doc/paper.tex
revised supplement. added some citations about core mixing

Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is an overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

\subsubsection{Covariate transforms and standardizations}

Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges. This standardization means that a regression coefficient associated with this covariate describes the change in extinction probability per change in standard deviation of geographic range, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}. Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.

@@

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla. The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each Coefficients for each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.

>>>>

The estimated age of each occurrence is based on the core-specific age-model that observation is from and can be overly precise. To alleviate this overprecision, we coarsened our temporal information in an effort to limit the effects of between-core heterogeneity in age. The occurrence histories of each species was then summarized as a series of binary codes indicating the presence or last occurrence of that species. For every occurrence of a species, except the last, that species existence and survival is recorded as a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is recorded as 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is an overconfident statement of preservation and may be shortening the actual durations of the studied species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang2016}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores \citep{Mekik2018,Broecker1999,Mekik2014,Peng1984}.

\subsubsection{Covariate transformation and standardization}

Prior to analysis, geographic range was then log-plus-one transformed and standardized by mean-centering the data and then dividing by the standard deviation of the distribution of geographic ranges. This standardization means that a regression coefficient associated with each covariate describes the change in extinction probability per change in standard deviation of that covariate, that coefficients associated with similarly standardized covariates will be directly comparable in magnitude, and that the intercept term corresponds to the expected value of the outcome at when geographic range is its average value \citep{ARM}. Change in geographic range between observations was measured from the standardized geographic range values and was not standardized separately.

@@

The matrix of time and phylum varying regression coefficients describing the effects of the covariates on a species' risk of extinction is called \(B\) -- a  \(w\) by \(p\) matrix, were \(w\) is the number of time temporal intervals and \(p\) is the number of phyla. The elements of this matrix, the regression coefficients, are themselves modeled as being multivariate normally distributed with vector of means \(\alpha\) describing the average intercept and regression coefficient estimates of each phylum \(p\). These phylum averages are themselves modeled as multivariate normally distributed with mean vector \(\mu\) describing the overall average regression coefficients, including the intercept. \(\mu\) has length \(D\) and is ordered intercept, range coefficient, change in range coefficient, temperature coefficient, temperature lag coefficient.
